%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% acknowledgements.tex                                                       %
% This file formats and includes the content of the acknowledgements section %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\cleardoublepage
\thispagestyle{empty}

\chapter{Acknowledgements}
\noindent The creation of this thesis involved significant support from many
throughout this journey. While lots have undoubtedly helped me grow
both academically and personally, there are a few I would like to express my 
sincere gratitude to. \newline

\noindent First, to my thesis advisor, Dr. Melton. His subject matter expertise is obvious, yet, 
perhaps more importantly, so is his passion for teaching others. Whether through classes or this research,
Dr. Melton consistently strives for the best for his students. Dr. Melton has been a beacon in my academic 
and personal undergraduate experience and I cannot thank him enough. \newline

\noindent Second, to my mother Brenda and sister Lindsey. You two have unwaveringly supported me 
in every one of my undertakings for as long as I can remember. 
You have pushed me, loved me, and have helped shape me into the person I am today.
You two are my rocks. I love you both endlessly.\newline

\noindent Lastly, to my friends. You make every day enjoyable and have shown me what happiness looks and
feels like. I am so thankful to have you by my side every day; you know I will always be by yours.
\cleardoublepage
